---
title: "Downstream analysis on scRNA-seq from mouse spermatogenesis using SCALPEL"
execute:
    cache: true
    fig-width: 8
    fig-height: 5
format:
    html:
        toc: true
        toc-location: right
        toc-expand: true
        toc-depth: 5
        number-sections: true
        code-fold: false
        theme: cosmo
        css: styles.css
---

```{r libraries_imports, cache=T, output=F, include=T}

library(Seurat)
library(dplyr)
library(data.table)
library(stringr)
library(Gviz)
library(ggplot2)
library(ggthemes)
library(rtracklayer)

#load gene based analysis Seurat object
dge.obj = readRDS("./mouse_DGEobj.RDS")
```


```{r custom_funcs, echo=F, cache=T}

#custom function for UMAP plots
theme_umap = function(ggobj) {
    axis <- ggh4x::guide_axis_truncated(
    trunc_lower = unit(0, "npc"),
    trunc_upper = unit(2, "cm"))

    ggobj +
    xlab("UMAP1") + 
    ylab("UMAP2") + 
    guides(x = axis, y = axis) + 
    theme(axis.line = element_line(arrow = arrow(), size=1), axis.title = element_text(hjust = 0)) + 
    scale_x_continuous(breaks = NULL) + scale_y_continuous(breaks = NULL)
}
```


::: {.callout-note}
**Note**: Install scalpelR for downstream analysis
:::

For execution of several function used in this downstream analysis with SCALPEL, we will implemented the R package scalpelR that can be installed from its [Github page](https://github.com/Franzx7/scalpelR) as:

```{r scalpelR_install, eval=F, include=T, cache=T}
install.packages("devtools")
devtools::install_github("Franzx7/scalpelR")

library(scalpelR)
```

## Introduction

In this example, we performed both gene-level and isoform-level single-cell analysis on a 10X 3' Chromium scRNA-seq dataset from mouse spermatogenesis. Using Seurat, we first conducted a standard single-cell gene-level analysis, which enabled the identification of distinct cell populations.

```{r gene_analysis, warning=F, include=T, cache=T}
#| fig-width: 7
#| fig-height: 4.5

#Visualize gene based sc cells
Seurat::DimPlot(dge.obj, label=T, repel = T, pt.size = 1., label.size = 6, group.by="ClustersL") %>%
 theme_umap() + 
 ggtitle("DGE based clusters")
```

Prior to obtaining these final clusters of populations, we performed some quality filtering steps to obtain this set of reliable cells. Using those same cell barcode tags, we executed SCALPEL on this 10X 3' Chromium scRNA-seq dataset. All the information required to run SCALPEL can be found on its [Github page](https://github.com/plasslab/SCALPEL).

::: {.callout-note}
**Note**: SCALPEL requires the FASTQ and cellranger repository associated to each sample for its execution. For 10x Genomics experiments, the repository can be obtained using the [CellRanger](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/what-is-cell-ranger) pipeline.

SCALPEL execution can be performed on a specific cell population using the **_- -barcodes_** argument option.
:::


Following SCALPEL execution, a result folder is generated that contains all the information required to perform downstream analysis.

```{rmd scalpel_execution_results, cache=T, include=T}
> tree -L 1 scalpel_output
scalpel_output
├── DIU_table.csv
├── iDGE_seurat.RDS
├── Runfiles
├── SRR6129050_APADGE.txt
├── SRR6129050_filtered.bam
├── SRR6129050_filtered.bam.bai
├── SRR6129050_seurat.RDS
├── SRR6129051_APADGE.txt
├── SRR6129051_filtered.bam
├── SRR6129051_filtered.bam.bai
└── SRR6129051_seurat.RDS

2 directories, 10 files
```

## Downstream analysis of SCALPEL results

### Loading isoform based Seurat object

Let's load the isoform-based Seurat object generated by SCALPEL using the reliable cell barcodes obtained from the gene-based analysis:
```{r load_idge_seurat, cache=T, include=T, echo=F, eval=T, results="hide"}
idge.obj = readRDS("./mouse_iDGE.RDS")
```

```{r load_idge_seurat2, cache=T, include=T, echo=T, eval=F}
idge.obj = readRDS("./iDGE_seurat.RDS")
```

### Quality filtering on low expressed isoforms features

For this analysis, we already used a reliable set of cells obtained from the gene-based analysis. However, we can still perform some filtering steps to discard lowly expressed isoforms features that are not informative. For example, we can filter out isoforms that are detected in less than 3 cells or more:
```{r quality_filtering, cache=T, include=T, warning=F, eval=F}

# Filter out isoforms that are detected in less than 3 cells
idge.obj = subset(idge.obj, features = rownames(idge.obj)[Matrix::rowSums(idge.obj@assays$RNA@counts > 0) >= 3])
```

### Dimensionality reduction and clustering

Let's perform the dimensionality reduction and clustering on this isoform-based Seurat object:
```{r load_isoform_seurat, cache=T, include=T, eval=F}

# Standard workflow for dimensionality reduction and clustering (Adjust the parameters)
idge.obj = NormalizeData(idge.obj)
# as isoform data is more sparse and noisy, we recommend to use a higher number of variable features
idge.obj = FindVariableFeatures(idge.obj, nfeatures = 6000) 
idge.obj = ScaleData(idge.obj)
idge.obj = RunPCA(idge.obj, npcs = 50)
idge.obj = RunUMAP(idge.obj, dims = 1:9)

#Clustering
idge.obj = FindNeighbors(idge.obj, dims = 1:9)
idge.obj = FindClusters(idge.obj, resolution = 0.9)
idge.obj$ClustersL = idge.obj$seurat_clusters

Seurat::DimPlot(idge.obj, label=T, repel = T, pt.size = 1., label.size = 6, group.by="ClustersL") %>%
 theme_umap() + 
 ggtitle("iDGE based clusters")
```

```{r idge_umap, cache=T, include=T, eval=T, warning=F, echo=F}
#| fig-width: 7
#| fig-height: 4.5

Seurat::DimPlot(idge.obj, label=T, repel = T, pt.size = 1., label.size = 6, group.by="ClustersL") %>%
 theme_umap() + 
 ggtitle("iDGE based clusters")
```


### Differential isoform expression analysis

We can perform a differential isoform expression analysis between the different clusters obtained from the isoform-based analysis. For example, let's perform a differential isoform expression analysis between the iPSC and NP cell populations:

```{r diu_analysis2, cache=T, include=T, warning=F, echo=T, eval=T}
diu.table = scalpelR::FindIsoforms(
    seurat_obj = idge.obj,
    group_by = "ClustersL",
    assay="RNA",
    threshold_pval = 0.05)

print(diu.table)
```

We can visualize a particular gene expression, for example the gene **Arl2bp**, that have differentially expressed isoforms between the iPSC and NPC populations:
```{r genec_tab, cache=T, include=T, warning=F}

gene.tab = diu.table %>% dplyr::filter(gene_name == "Arl2bp")
print(gene.tab)
```

### Visualization of isoform expression in cells

#### UMAP

We can visualize the expression of these differentially expressed isoforms in the cells using a UMAP plot using Seurat. For example, let's visualize the expression of the two differentially expressed isoforms of the gene **Arl2bp**:

```{r genec_umap, cache=T, include=T, warning=F}
#| fig-width: 8
#| fig-height: 3.5

Idents(idge.obj) = idge.obj$ClustersL
Seurat::FeaturePlot(idge.obj,
   features = gene.tab$gene_tr,
   pt.size = 1,
   label.size=6,
   order=T,
   label=T,
   repel=T,
   cols = c("gray80", "red"))
```

#### Violin plots

Aditionally, we can visualize the expression of these differentially expressed isoforms within the condition using violin plot as:
```{r arl2bp_vln1, cache=T, include=T, warning=F, eval=T}
#| fig-width: 6
#| fig-height: 10

p1 = scalpelR::plotExpression(
    seurat_obj = idge.obj,
    features = gene.tab$gene_tr,
    group_var = "ClustersL",
    plot = "group") +
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

p2 = scalpelR::plotExpression(
    seurat_obj = idge.obj,
    features = gene.tab$gene_tr,
    group_var = "ClustersL",
    plot = "features") +
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

list(p1, p2) %>% patchwork::wrap_plots(ncol=1)
```

### Visualization of read coverage on the genome

Furthermore, we can visualize the read coverage of these differentially expressed isoforms using the BAM files generated by SCALPEL. For example, let's visualize the read coverage of the two differentially expressed isoforms of the gene **Arl2bp**.

As we used a curated set of cell barcodes from the gene-based analysis to run SCALPEL, the BAM files generated by SCALPEL already contains only the reads from those reliable cells. Therefore, we can directly use these filtered BAM files to visualize the read coverage of the differentially expressed isoforms. However, if you want to visualize the read coverage of a specific cell population, you can filter the BAM files using the cell barcodes associated to that population and scalpelR subsetBAM as:
```{r plot_coverage, cache=T, include=T, eval=F}

es_barcodes = dplyr::filter(idge.obj[[]], ClustersL=="elongated spermatids") %>% rownames()
rs_barcodes = dplyr::filter(idge.obj[[]], ClustersL=="round spermatids") %>% rownames()
sc_barcodes = dplyr::filter(idge.obj[[]], ClustersL=="spermatocytes") %>% rownames()

#let's merge the SRR6129050 and SRR6129051 BAM files generated by SCALPEL
system("samtools merge -o SRR612905X.bam SRR6129050_filtered.bam SRR6129051_filtered.bam -@ 4")

# Filter BAM file to obtain reads from a specific cell population
#elongated spermatids population
scalpelR::subsetBAM(
    bam_file = "./SRR612905X.bam",
    barcodes = es_barcodes,
    output_file = "./subset_es_filtered.bam",
    tag = "CB",
    cores = 4)

#round spermatids population
scalpelR::subsetBAM(
    bam_file = "./SRR612905X.bam",
    barcodes = rs_barcodes,
    output_file = "./subset_rs_filtered.bam",
    tag = "CB",
    cores = 4)

#spermatocytes population
scalpelR::subsetBAM(
    bam_file = "./SRR612905X.bam",
    barcodes = sc_barcodes,
    output_file = "./subset_sc_filtered.bam",
    tag = "CB",
    cores = 4)

#Plot coverage using the subset BAM files and GTF annotation reference
gtrack = rtracklayer::import("./gencode.vM21.annotation.gtf.gz")
scalpelR::CoveragePlot(
    genome_gr = gtrack,
    bamfiles = c("ES"="./subset_es_filtered.bam",
                 "RS"="./subset_rs_filtered.bam",
                 "SC"="./subset_rs_filtered.bam"),
    gene_in = "Arl2bp",
    genome_sp = "mm10")
```

```{r plot_coverage2, cache=T, include=T, warning=F, echo=F}
#| fig-width: 8
#| fig-height: 5

gtrack = rtracklayer::import("./gencode.vM21.annotation.gtf.gz")
scalpelR::CoveragePlot(
    genome_gr = gtrack,
    bamfiles = c("SC"="./subset_sc_filtered.bam",
                 "RS"="./subset_rs_filtered.bam",
                 "ES"="./subset_es_filtered.bam"),
    features = c("Arl2bp-201", "Arl2bp-204"),
    sizes = c(358,1189,495),
    gene_in = "Arl2bp",
    genome_sp = "mm10")
```

By default, the CoveragePlot function will plot all the isoforms of the gene of interest. However, you can also specify a set of isoforms to plot using the **features** argument option. Furthermore, you can also specify the distance to zoom from the 3'UTR using the **distZOOM** argument option as:

```{r distzoom, cache=T, include=T, warning=F, echo=F}
#| fig-width: 8
#| fig-height: 5
#| 
scalpelR::CoveragePlot(
    genome_gr = gtrack,
    bamfiles = c("SC"="./subset_sc_filtered.bam",
                 "RS"="./subset_rs_filtered.bam",
                 "ES"="./subset_es_filtered.bam"),
    features = c("Arl2bp-201", "Arl2bp-204"),
    sizes = c(358,1189,495),
    gene_in = "Arl2bp",
    genome_sp = "mm10",
    distZOOM = 2000)
```

### 3'UTR length analysis in single-cells

Using the isoform quantification obtained from SCALPEL, and the genome annotation reference, we can estimate the 3'UTR length of each isoform in each cell. For this, we can use the function **scalpelR::weightedUTR_cell** as:

```{r 3utr_length, cache=T, include=T, eval=T, echo=T}

utr3length_tab = scalpelR::weightedUTR_cell(
    seurat_obj = idge.obj,
    annotation_gr = gtrack,
    group.by = "ClustersL",
    assay = "RNA") %>%
    dplyr::mutate(group=factor(group,
     levels=c("spermatocytes", "round spermatids", "elongated spermatids")))

print(utr3length_tab %>% head(5))
```

Then, based on this metric, we can visualize the 3'UTR length distribution in the different cell populations using plots as:
```{r 3utr_plots, cache=T, include=T, warning=F, eval=T}
#| fig-width: 7
#| fig-height: 6

utr3length_tab %>%
    ggplot2::ggplot(aes(x=group, y=wUTRlength.cell, fill=group)) +
    geom_jitter(na.rm=T, size=.5) +
    ggplot2::geom_violin(scale="width", trim=T, na.rm = T, alpha=.6) +
    ggplot2::geom_boxplot(fill="white", width=.3) +
    ggplot2::ylim(c(0,700)) +
    ggplot2::ggtitle("average 3'UTR length distribution per cell") +
    ggplot2::theme(title = element_text(size=9)) +
    ggplot2::theme_bw(base_size = 13)
```

## Further downstream analysis

Using the R Seurat Object generated by SCALPEL, you can perform further downstream analysis in the same way as a standard gene-based Seurat object. For example, you can perform trajectory analysis using Monocle3, etc... \
For more information, please refer to the Seurat documentation [here](https://satijalab.org/seurat/).