---
title: "Downstream analysis on scRNA-seq from human neuronal differentiation using SCALPEL"
execute:
    cache: true
    fig-width: 8
    fig-height: 5
format:
    html:
        toc: true
        toc-location: right
        toc-expand: true
        toc-depth: 5
        number-sections: true
        code-fold: false
        theme: cosmo
        css: styles.css
---

```{r libraries_imports, cache=T, output=F, include=T}

library(Seurat)
library(dplyr)
library(data.table)
library(stringr)
library(Gviz)
library(ggplot2)
library(ggthemes)
library(rtracklayer)

#load gene based analysis Seurat object
dge.obj = readRDS("./neuro_DGE.RDS")
```


```{r custom_funcs, echo=F, cache=T}

#custom function for UMAP plots
theme_umap = function(ggobj) {
    axis <- ggh4x::guide_axis_truncated(
    trunc_lower = unit(0, "npc"),
    trunc_upper = unit(2, "cm"))

    ggobj +
    xlab("UMAP1") + 
    ylab("UMAP2") + 
    guides(x = axis, y = axis) + 
    theme(axis.line = element_line(arrow = arrow(), size=1), axis.title = element_text(hjust = 0)) + 
    scale_x_continuous(breaks = NULL) + scale_y_continuous(breaks = NULL)
}
```

::: {.callout-note}
**Note**: Install scalpelR for downstream analysis
:::

For execution of several function used in this downstream analysis with SCALPEL, we will implemented the R package scalpelR that can be installed from its [Github page](https://github.com/Franzx7/scalpelR) as:

```{r scalpelR_install, eval=F, include=T, cache=T}
install.packages("devtools")
devtools::install_github("Franzx7/scalpelR")

library(scalpelR)
```

## Introduction

In this exemple, we did a gene and isoform based single-cell analysis from this Drop-seq single-cell dataset. Using Seurat, we first performed a standard single-cell gene-level analysis of the human neuronal differentiation dataset, which enabled the identification of distinct cell populations.

```{r gene_analysis, warning=F, include=T, cache=T}
#| fig-width: 6
#| fig-height: 3.5

#Visualize gene based sc cells
DimPlot(dge.obj, label=T, repel = T, pt.size = 1., label.size = 6) %>% theme_umap() +
    ggtitle("iPS & NPC cells clusters")
```

Prior to obtain this final clusters of population, we performed some quality filtering steps to obtain this set of reliable cells. Using those same cells barcode tags, we executed SCALPEL on this Drop-seq based single-cell dataset. All the information required to run SCALPEL can be found on its [Github page](https://github.com/plasslab/SCALPEL).

::: {.callout-note}
**Note**: SCALPEL requires the FASTQ, BAM and Gene count expression associated to each sample for its execution. For Drop-seq experiments, the BAM and Gene count expression can be obtained using the [Drop-seq tools](http://mccarrolllab.com/dropseq/) and for 10x Genomics experiments, the BAM and Gene count expression can be obtained using the [CellRanger](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/what-is-cell-ranger) pipeline.

SCALPEL execution can be performed on a specific cell population using the **_- -barcodes_** argument option.
:::


Following SCALPEL execution, a result folder is generated that contains all the information required to perform downstream analysis.

```{rmd scalpel_execution_results, cache=T, include=T}
> tree -L 1 scalpel_output
scalpel_output
├── DIU_table.csv
├── hiPSCs_APADGE.txt
├── hiPSCs_filtered.bam
├── hiPSCs_filtered.bam.bai
├── hiPSCs_seurat.RDS
├── iDGE_seurat.RDS
├── NPCs_APADGE.txt
├── NPCs_filtered.bam
├── NPCs_filtered.bam.bai
├── NPCs_seurat.RDS
└── Runfiles

2 directories, 10 files
```

## Downstream analysis of SCALPEL results

### Loading isoform based Seurat object

Let's load the isoform-based Seurat object generated by SCALPEL using the reliable cell barcodes obtained from the gene-based analysis:
```{r load_idge_seurat, cache=T, include=T, echo=T}
idge.obj = readRDS("./iDGE_seurat.RDS")
idge.obj
```

### Quality filtering on low expressed isoforms features

For this analysis, we already used a reliable set of cells obtained from the gene-based analysis. However, we can still perform some filtering steps to discard lowly expressed isoforms features that are not informative. For example, we can filter out isoforms that are detected in less than 3 cells:
```{r quality_filtering, cache=T, include=T, warning=F, eval=F}

# Filter out isoforms that are detected in less than 3 cells
idge.obj = subset(idge.obj, features = rownames(idge.obj)[Matrix::rowSums(idge.obj@assays$RNA@counts > 0) >= 3])
```

### Dimensionality reduction and clustering

Let's perform the dimensionality reduction and clustering on this isoform-based Seurat object:
```{r load_isoform_seurat, cache=T, include=T, eval=F}

# Standard workflow for dimensionality reduction and clustering (Adjust the parameters)
idge.obj = NormalizeData(idge.obj)
# as isoform data is more sparse and noisy, we recommend to use a higher number of variable features
idge.obj = FindVariableFeatures(idge.obj, nfeatures = 4000) 
idge.obj = ScaleData(idge.obj)
idge.obj = RunPCA(idge.obj, npcs = 50)
idge.obj = RunUMAP(idge.obj, dims = 1:30)

#Clustering
idge.obj = FindNeighbors(idge.obj, dims = 1:30)
idge.obj = FindClusters(idge.obj, resolution = 0.5)
idge.obj$Clusters = idge.obj$seurat_clusters

```

Now, we can visualize the clusters obtained from this isoform-based analysis. We can observe that while briging a new layer of information, the isoform quantification conserves the cell type specificity identified from the gene-based analysis
```{r isoform_clusters, cache=T, include=T, warning=F}
#| fig-width: 6
#| fig-height: 3.5

# Filter out isoforms that are detected in less than 3 cells
DimPlot(idge.obj, group.by="Clusters", label=T, repel = T, pt.size = 1., label.size = 6) %>% theme_umap() + ggtitle("iPS & NPC cells clusters (isoform-based)")
```

### Differential isoform expression analysis

We can perform a differential isoform expression analysis between the different clusters obtained from the isoform-based analysis. For example, let's perform a differential isoform expression analysis between the iPSC and NP cell populations:

```{r diu_analysis, cache=T, include=T, eval=F}

diu.table = scalpelR::FindIsoforms(
    seurat_obj = idge.obj,
    group_by = "Clusters",
    assay="RNA",
    threshold_pval = 0.05)

print(diu.table)
```
Aggregating counts in conditions...
Performing filtering and comparison test...
  |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=51s 

```{r diu_analysis2, cache=T, include=F, warning=F, echo=F, eval=T, results='hide'}
diu.table = scalpelR::FindIsoforms(
    seurat_obj = idge.obj,
    group_by = "Clusters",
    assay="RNA",
    threshold_pval = 0.05)
```

```{r print_diu,  cache=T, include=T, warning=F, echo=F}
print(diu.table)
```

We can visualize a particular gene expression, for example the gene **EIF1**, that have differentially expressed isoforms between the iPSC and NPC populations:
```{r eif1_tab, cache=T, include=T, warning=F}

gene.tab = diu.table %>% dplyr::filter(gene_name == "EIF1")
print(gene.tab)
```

### Visualization of isoform expression in cells

#### UMAP

We can visualize the expression of these differentially expressed isoforms in the cells using a UMAP plot using Seurat. For example, let's visualize the expression of the two differentially expressed isoforms of the gene **EIF1**:

```{r eif1_umap, cache=T, include=T, warning=F}
#| fig-width: 8
#| fig-height: 3.5

Seurat::FeaturePlot(idge.obj,
    features = gene.tab$gene_tr,
    pt.size = 1, 
    label.size=6, 
    order=T, 
    label=T, 
    cols = c("gray80", "red"))
```

#### Violin plots

Aditionally, we can visualize the expression of these differentially expressed isoforms within the condition using violin plot as:
```{r eif1_vln1, cache=T, include=T, warning=F, eval=F, results='hide'}
#| fig-width: 6
#| fig-height: 8

p1 = scalpelR::plotExpression(
    seurat_obj = idge.obj,
    features = gene.tab$gene_tr,
    group_var = "Clusters",
    plot = "group")

p2 = scalpelR::plotExpression(
    seurat_obj = idge.obj,
    features = gene.tab$gene_tr,
    group_var = "Clusters",
    plot = "features")

list(p1, p2) %>% patchwork::wrap_plots(ncol=1)
```

```{r eif1_vln2, cache=T, include=T, warning=F, echo=F}
#| fig-width: 6
#| fig-height: 8

p1 = scalpelR::plotExpression(
    seurat_obj = idge.obj,
    features = gene.tab$gene_tr,
    group_var = "Clusters",
    plot = "group")

p2 = scalpelR::plotExpression(
    seurat_obj = idge.obj,
    features = gene.tab$gene_tr,
    group_var = "Clusters",
    plot = "features")

list(p1, p2) %>% patchwork::wrap_plots(ncol=1)
```


### Visualization of read coverage on the genome

Furthermore, we can visualize the read coverage of these differentially expressed isoforms using the BAM files generated by SCALPEL. For example, let's visualize the read coverage of the two differentially expressed isoforms of the gene **EIF1**.

As we used a curated set of cell barcodes from the gene-based analysis to run SCALPEL, the BAM files generated by SCALPEL already contains only the reads from those reliable cells. Therefore, we can directly use these filtered BAM files to visualize the read coverage of the differentially expressed isoforms. However, if you want to visualize the read coverage of a specific cell population, you can filter the BAM files using the cell barcodes associated to that population and scalpelR subsetBAM as:
```{r plot_coverage, cache=T, include=T, eval=F}

ips_barcodes = dplyr::filter(idge.obj[[]], Clusters=="iPS") %>% rownames()
np_barcodes = dplyr::filter(idge.obj[[]], Clusters=="NP") %>% rownames()

# Filter BAM file to obtain reads from a specific cell population
#IPS population
scalpelR::subsetBAM(
    bam_file = "./scalpel_output/hiPSCs_filtered.bam",
    barcodes = ips_barcodes,
    output_file = "./subset_ips_filtered.bam",
    tag = "XC",
    cores = 4)

#NP population
scalpelR::subsetBAM(
    bam_file = "./scalpel_output/NPCs_filtered.bam",
    barcodes = ips_barcodes,
    output_file = "./subset_np_filtered.bam",
    tag = "XC",
    cores = 4)

#Plot coverage using the subset BAM files and GTF annotation reference
gtrack = rtracklayer::import("./gencode.v41.annotation.gtf.gz")
scalpelR::CoveragePlot(
    genome_gr = gtrack,
    bamfiles = c("IPS"="./subset_ips_filtered.bam",
                 "NP"="./subset_np_filtered.bam"),
    gene_in = "EIF1",
    genome_sp = "hg38")
```


```{r plot_coverage2, cache=T, include=T, warning=F, echo=F}
#| fig-width: 8
#| fig-height: 4

gtrack = rtracklayer::import("~/Projects/scalpel_wiki2/datas/gencode.v41.annotation.gtf.gz")
scalpelR::CoveragePlot(
    genome_gr = gtrack,
    bamfiles = c("IPS"="~/Projects/iPS_filtered.bam",
                 "NP"="~/Projects/NP_filtered.bam"),
    gene_in = "EIF1",
    genome_sp = "hg38")
```

By default, the CoveragePlot function will plot all the isoforms of the gene of interest. However, you can also specify a set of isoforms to plot using the **features** argument option. Furthermore, you can also specify the distance to zoom from the 3'UTR using the **distZOOM** argument option as:

```{r dist_zoom, cache=T, include=T, warning=F, echo=T}
#| fig-width: 8
#| fig-height: 4

scalpelR::CoveragePlot(
    genome_gr = gtrack,
    bamfiles = c("IPS"="~/Projects/iPS_filtered.bam",
                 "NP"="~/Projects/NP_filtered.bam"),
    features = gene.tab$transcript_name,
    gene_in = "EIF1",
    genome_sp = "hg38",
    distZOOM = 2000)
```


### 3'UTR length analysis in single-cells

Using the isoform quantification obtained from SCALPEL, and the genome annotation reference, we can estimate the 3'UTR length of each isoform in each cell. For this, we can use the function **scalpelR::weightedUTR_cell** as:

```{r 3utr_length, cache=T, include=T, eval=T}

utr3length_tab = scalpelR::weightedUTR_cell(
    seurat_obj = idge.obj,
    annotation_gr = gtrack,
    group.by = "Clusters",
    assay = "RNA"
)

print(utr3length_tab)
```

Then, based on this metric, we can visualize the 3'UTR length distribution in the different cell populations using plots as:
```{r 3utr_plots, cache=T, include=T, warning=F, eval=T}
#| fig-width: 4
#| fig-height: 5

utr3length_tab %>%
    ggplot2::ggplot(aes(x=group, y=wUTRlength.cell, fill=group)) +
    geom_jitter(na.rm=T, size=.5) +
    ggplot2::geom_violin(scale="width", trim=T, na.rm = T, alpha=.6) +
    ggplot2::geom_boxplot(fill="white", width=.3) +
    ggplot2::ylim(c(0,1000)) +
    ggplot2::ggtitle("average 3'UTR length distribution per cell") +
    ggplot2::theme(title = element_text(size=9)) +
    ggplot2::theme_bw(base_size = 9)
```

## Further downstream analysis

Using the R Seurat Object generated by SCALPEL, you can perform further downstream analysis in the same way as a standard gene-based Seurat object. For example, you can perform trajectory analysis using Monocle3, etc... \
For more information, please refer to the Seurat documentation [here](https://satijalab.org/seurat/).